Formalizing the Edmonds-Karp Algorithm

### 2016-03-02 paper/root.tex
Added TODOs to latex file

This paper present a formalization of the correctness proof of the Ford-Fulkerson method in Isabelle/HOL~\cite{NPW02}. Our formalization is based on the informal proof presented in the book \emph{Introduction to algorithms}~\cite{CLRS09}. We then provide a verified implementation of the Edmons-Karp's algorithm~\cite{EK72}. This algorithm instantiates the Ford-Fulkerson method with breadth-first-search and provides a polynomial algorithm for solving the maximum flow problem in generic case--- maximum flow on networks with real edge capacities.

@@

One solution for the aforementioned problem is refinement~\cite{Wirth71}. In order to generate executable code of an algorithm using refinement, we first formulate the algorithm on an abstract level. The abstract version of the algorithm has a clean correctness proof as it only captures the idea behind the algorithm. Next, we refine the abstract definition of the algorithm towards an executable implementation in possibly multiple refinement steps. During each step, we only need to prove the correctness of the implementation of a particular abstract concept. Hence, we can be sure about the correctness of the resulting executable program as each refinement step preserves the correctness.

There are several approaches to data refinement in Isabelle/HOL. We will be using the Isabelle Refinement Framework~\cite{LaTu12,La12}, which has been used for proving more complex results such as formalized implementation of Hopcroft’s DFA-minimization algorithm~\cite{LaTu12}. Given an algorithm phrased over abstract concepts like sets and maps, it automatically synthesizes a concrete, executable algorithm and the corresponding refinement theorem. To make it applicable for the development of actual algorithms, Autoref is integrated with the Isabelle Refinement Framework~\cite{La12} and the Isabelle Collection Framework~\cite{LL10}.

@@

There are two kinds of edges in the residual graph: 1) those of the residual graph that present the edges of the flow network with left-over capacities; and 2) those of the residual graph that do not occur in the flow network, and indicate the possibility for pushing the flow backward. Formally speaking, consider a flow $f$ on an $s$-$t$ network $G = (V, E)$ with the capacity function $c$. A residual capacity function $c_f \colon V \times V \rightarrow \mathbb{R}$ is defined as following:
\[ c_f (u, v) = 
	\begin{cases}
	c (u, v) - f(u, v) \hfill & \text{ if $(u, v) \in E$} \\
	f (v, u) \hfill & \text{ if $(v, u) \in E$} \\	
	0 \hfill & \text{ otherwise} \\
	\end{cases} 
\]

@@

The Ford-Fulkerson algorithm is sometimes called a method instead of an algorithm because it does not fully specify the procedure for finding the augmenting path in the residual graph. There are several implementations with different execution times. For instance, the Edmond-Karp algorithm uses breadth-first-search (BFS) for finding augmenting paths and has $O (VE^2)$ running-time. The augmentings paths that are computed using BFS are also shortest paths connecting source to the sink in the residual graph. Edmond and Karp showed that in the Ford-Fulkerson algorithm, if each augmenting path is the shortest one, the length of the augmenting paths is non-decreasing and the algorithm always terminates. Using this statement they could finally present a polynomial algorithm for solving the maximum flow problem in generic case of real edge capacities.

@@

Informal proofs focus on the relevant thoughts by leaving out technical details. However, in a formal proof each step has to be precisely specified as a valid application of some inference rule which makes formal proofs often lengthy and hard to follow. In the case that an interactive theorem prover is used, readers may even need to replay the proof to understand the idea behind it. In Isabelle proof assistant, the Isar proof language bridges the semantic gap between internal notions of proof given by interactive theorem proving systems and an appropriate level of abstraction for user-level work. Isar makes formal proofs readable even for those who are not familiar with Isabelle concepts. We have used such a functionality to present our formal proof of the Ford-Fulkerson method which resembels the informal proof described by Cormen et al.~\cite{CLRS09}.

@@

  We observe that, for sparse graphs, the Java implementation is roughly faster by a factor of 6, while for dense graphs, our implementation comes quite close 
  to the Java implementation. We have no full explanation for this phenomenon. Note that the Java implementation operates on flows, while our implementation 
  operates on residual graphs (\cf Section~\ref{sec:impl_res_graph}). Moreover, the Java implementation does not store the augmenting 
  path in an intermediate list, but uses the predecessor map computed by the BFS directly (\cf Section~\ref{sec:impl_datastructures}).

>>>>

(TODO: Here I have mentioned the Edmonds-Karp for the first time.) This paper present a formalization of the correctness proof of the Ford-Fulkerson method in Isabelle/HOL~\cite{NPW02}. Our formalization is based on the informal proof presented in the book \emph{Introduction to algorithms}~\cite{CLRS09}. We then provide a verified implementation of the Edmons-Karp's algorithm~\cite{EK72}. This algorithm instantiates the Ford-Fulkerson method with breadth-first-search and provides a polynomial algorithm for solving the maximum flow problem in generic case--- maximum flow on networks with real edge capacities.

@@

(TODO: I am not sure about the level of details that I describe refinement in introduction, co-relate this with the refinement section) One solution for the aforementioned problem is refinement~\cite{Wirth71}. In order to generate executable code of an algorithm using refinement, we first formulate the algorithm on an abstract level. The abstract version of the algorithm has a clean correctness proof as it only captures the idea behind the algorithm. Next, we refine the abstract definition of the algorithm towards an executable implementation in possibly multiple refinement steps. During each step, we only need to prove the correctness of the implementation of a particular abstract concept. Hence, we can be sure about the correctness of the resulting executable program as each refinement step preserves the correctness.

(TODO: check the citation order and description) There are several approaches to data refinement in Isabelle/HOL. We will be using the Isabelle Refinement Framework~\cite{LaTu12,La12}, which has been used for proving more complex results such as formalized implementation of Hopcroft’s DFA-minimization algorithm~\cite{LaTu12}. Given an algorithm phrased over abstract concepts like sets and maps, it automatically synthesizes a concrete, executable algorithm and the corresponding refinement theorem. To make it applicable for the development of actual algorithms, Autoref is integrated with the Isabelle Refinement Framework~\cite{La12} and the Isabelle Collection Framework~\cite{LL10}.

@@

(TODO: there is a shorter informal description available as comment. Adapt that one if required) There are two kinds of edges in the residual graph: 1) those of the residual graph that present the edges of the flow network with left-over capacities; and 2) those of the residual graph that do not occur in the flow network, and indicate the possibility for pushing the flow backward. Formally speaking, consider a flow $f$ on an $s$-$t$ network $G = (V, E)$ with the capacity function $c$. A residual capacity function $c_f \colon V \times V \rightarrow \mathbb{R}$ is defined as following:
\[ c_f (u, v) = 
	\begin{cases}
	c (u, v) - f(u, v) \hfill & \text{ if $(u, v) \in E$} \\
	f (v, u) \hfill & \text{ if $(v, u) \in E$} \\	
	0 \hfill & \text{ otherwise} \\
	\end{cases} 
\]

@@

(TODO: Here I mention the Edmonds-Karp algorithm and I mention its running time, co-relate this with definition in later sections) The Ford-Fulkerson algorithm is sometimes called a method instead of an algorithm because it does not fully specify the procedure for finding the augmenting path in the residual graph. There are several implementations with different execution times. For instance, the Edmond-Karp algorithm uses breadth-first-search (BFS) for finding augmenting paths and has $O (VE^2)$ running-time. The augmentings paths that are computed using BFS are also shortest paths connecting source to the sink in the residual graph. Edmond and Karp showed that in the Ford-Fulkerson algorithm, if each augmenting path is the shortest one, the length of the augmenting paths is non-decreasing and the algorithm always terminates. Using this statement they could finally present a polynomial algorithm for solving the maximum flow problem in generic case of real edge capacities.

@@

(TODO: not sure about repeating citation of Cormen) Informal proofs focus on the relevant thoughts by leaving out technical details. However, in a formal proof each step has to be precisely specified as a valid application of some inference rule which makes formal proofs often lengthy and hard to follow. In the case that an interactive theorem prover is used, readers may even need to replay the proof to understand the idea behind it. In Isabelle proof assistant, the Isar proof language bridges the semantic gap between internal notions of proof given by interactive theorem proving systems and an appropriate level of abstraction for user-level work. Isar makes formal proofs readable even for those who are not familiar with Isabelle concepts. We have used such a functionality to present our formal proof of the Ford-Fulkerson method which resembels the informal proof described by Cormen et al.~\cite{CLRS09}.

@@

  (TODO: Table's caption is incorrect, I think we have enough space to print the whole table with SML, Java as rows and graph size as columns. Also, single lines for separating table looks nicer. I have seen it in your Hopcroft's paper) We observe that, for sparse graphs, the Java implementation is roughly faster by a factor of 6, while for dense graphs, our implementation comes quite close 
  to the Java implementation. We have no full explanation for this phenomenon. Note that the Java implementation operates on flows, while our implementation 
  operates on residual graphs (\cf Section~\ref{sec:impl_res_graph}). Moreover, the Java implementation does not store the augmenting 
  path in an intermediate list, but uses the predecessor map computed by the BFS directly (\cf Section~\ref{sec:impl_datastructures}).
